\section{Automated and emerging technologies}
\subsection{Automated systems}

\textbf{1} \textit{Describe how sensors, microprocessors and
actuators can be used in collaboration to create
automated systems.}

An automated system consists of sensors, which measure the environment, a microprocessor, which
processes data from the sensors and determine if an action needs to be taken and actuators
to create a movement depending on the inputs gathered. 

Inputs gathered by the sensors are analogue, and must be converted to digital to be processed by
a microprocessor. This conversion is done by a digital to analogue converter or DAC. In the 
microprocessor, recieved values are compared against stored values, and depending on the range
of the values read by the sensor, signals are sent to actuators. The actuators perform analogue
actions, but recieve digital signals, which must be converted to analogue, which is done by means
of a digital to analogue converter or DAC. This process of taking readings through sensors,
passing it through the microprocessor and performing a corresponding actions repeats until the
system is switched off.

\bigskip

\noindent\textbf{2} \textit{Describe the advantages and disadvantages of an
automated system used for a given scenario.}

In almost all situations, initial costs for automated systems are high as development of the
technology and price of the technology itself is all quite expensive. The running costs of the
system is high as the system requires maintenance, but as fewer people need be employed, long
term costs will be lower. Absence of humans, and their distractions will mean its all very
safe, as robots will make less mistakes. These systems will replace human jobs, but jobs will
be created in maintenance of the system. These robots will work all day every day, with no 
complaints, increasin efficiency, and since robots make no mistakes, no errors will result.

\subsection{Robotics}
\noindent\textbf{1} \textit{Understand what is meant by robotics.}

Robotics is a branch of computer science that incorporates the design, construction and operation
of robots. Factory equipment, domestic robots and drones are examples.

\bigskip

\noindent\textbf{2} \textit{Describe the characteristics of a robot.}

A robot consists of a mechanical structure or framework; electrical components such as sensors,
microprocessors and actuators with functions the same as automated systems; and are programmable.

\bigskip

\noindent\textbf{3} \textit{Understand the roles that robots can perform
and describe the advantages and disadvantages
of their use}

Identical to Section 6.1.2.

\subsection{Artificial intelligence}
\noindent\textbf{1} \textit{Understand what is meant by artificial intelligence (AI).}

AI is a branch of computer science dealing with the simulation of intelligent behaviours of by
computers.

\bigskip

\noindent\textbf{2} \textit{Describe the main characteristics of AI as the
collection of data and the rules for using that
data, the ability to reason, and can include the
ability to learn and adapt.}

As given.

\bigskip

\noindent\textbf{3} \textit{Explain the basic operation and components of AI
systems to simulate intelligent behaviour.}

An expert system is a special type of system that attempts to emulate the expertise of a human.
It takes questions from a user and determines solutions or answers.

It consists of a knowledge base, a rule base, an inference engine and an user interface. The
knowledge base is a list of facts, the rule base links the facts. The inference engine is the
part of the program that decides what questions to ask next and what answers to give. The user
interface is that through which the user inputs questions and the program outputs questions and/or
solutions.

\medskip

Machine learning is when a program has the ability to automatically adapt its processes and/or 
data. It can be supervised or unsupervised. Supervised machine learning consists of a user tellin
what data being fed to it means. Unsupervised learning is when data is fed to the machine and the
machine learns from it.
