\section{The internet and its uses}
\subsection{The internet and the world wide web}

\noindent\textbf{1} \textit{Understand the difference between the internet and the world wide web.}

The internet is an infrastructure, it is a wide area network. The world wide web is the collective
term for all the websites and webpages available to access through the internet.

\bigskip

\noindent\textbf{2} \textit{Understand what is meant by a uniform resource locator (URL).}

A URL is a text based address for a webpage, it controls the protocol, domain name and 
webpage/webfile name. A URL has the following format:

\begin{center}
\begin{BVerbatim}
protocol://www.domain-name/webpage-name
\end{BVerbatim}
\end{center}

\bigskip

\noindent\textbf{3} \textit{Describe the purpose and operation of hypertext
transfer protocol (HTTP) and hypertext transfer
protocol secure (HTTPS).}

Webpages are written using the hypertext markup language (HTML), cascading style sheets (CSS) and
JavaScript. The hypertext transfer protocol, HTTP, is used for the transfer of these webpages
across the internet. HTTPS stands for hyptertext transfer protocol secure and adds a layer of
encryption and hence security to these transfers of webpages.

\bigskip

\noindent\textbf{4} \textit{Explain the purpose and functions of a web browser.}

The main purpose of a web browser is to render the HTML of webpages and hence display them. They
have several additional functions.

Modern web browsers allows users to bookmark certain webpages. The user's history is stored by
most web browsers and multiple tabs are provided allowing for simultaneous browsing. Navigation
tools amonsgst and through webpages, and an address bar is provided to the user, so that they can
enter the URL of the website they desire.

\newpage i
